Optogenetic Maxwell Demon to Exploit Intrinsic Noise and Control Cell Differentiation Despite Time Delays and Extrinsic Variability

### 2023-02-02 paper.tex
Edits up to section 3.1

In this paper, the analysis of this Single-Input-Multiple-Output (SIMO) multicellular control problem is extended to examine how the FAC and PAC controllers are affected by additional model uncertainties, including coarse grained approximations to the system dynamics, uncertainties or errors in system parameters, and delays in the biochemical sensors or optogenetic controllers through which the external feedback is implemented. In the next section, we briefly describe our methods to formulate the chemical master equation to analyze multiple individual cells in the same spatially homogeneous environment; we define the FAC and PAC controllers and their effects the dynamics of cells within that environment; and we explain how the control law is optimized to achieve desired performance objectives. Next, in the `Results' section, we explore how the granularity of the model formulation, the inaccuracy of parametric assumptions, and the inclusion of time delays would affect the FAC and PAC control performance. Finally, in `Conclusions' we summarize our findings and discuss the implications that these controllers, and their robustness to modeling errors, may have on future systems in synthetic biology.

@@

More specifically, the openloop infinitesimal generator, $\mathbf{A}_0$, is constructed according to:
\begin{equation}
[\mathbf{A}_0]_{ij} = \left\{
\begin{array}{rl}
-\sum_{\mu =1}^{2N_{\rm c}} w_{\mu}(\mathbf{X}_j), &\text{for }i=j,\\
w_{\mu}(\mathbf{X}_j), &\text{for }\mathbf{X}_i = \mathbf{X}_j+\mathbf{s}_\mu\\
0, & \text{otherwise,}
\end{array}\right. 
\label{InfA}
\end{equation}
and the feedback infinitesimal generator, $\mathbf{Bu}^{\mathcal{C}}$, of the controller is constructed according to
\begin{equation}
[\mathbf{Bu^{\mathcal{C}}]}_{ij} = \left\{
\begin{array}{rl}
- N_{\rm c}{u}^{\mathcal{C}}(\mathbf{X}_j), &\text{for }i=j\\
{u}^{\mathcal{C}}(\mathbf{X}_j), &\text{for }\begin{array}{ll}\mathbf{X}_i =\mathbf{X}_j + \mathbf{e}_{i_{\rm c}}, \\ \text{and } i_{\rm c} = 1,\ldots,N_{\rm c}\end{array},\\
0, & \text{otherwise,}
\end{array}\right.
\label{InfB}
\end{equation}
where ${u}^{\mathcal{C}}(\mathbf{X}_j)$ is the specification of the $\mathcal{C}$ controller in terms of the state, or its partial observations.

@@

\subsection{System Granularity Scalings}
One of the major computational obstacles in using the CME or FSP formulations to analyze discrete stochastic chemical kinetics is that the dimension of the FSP scales linearly with each chemical species, and therefore exponentially with the number of independent species. To circumvent this issue, several previous studies have projected the FSP onto lower dimensional spaces based using various techniques such as time scale separations\cite{Peles2006}, Krylov subspaces\cite{Macnemara&Sidje?}, coarse meshes\cite{Munsky2008IEEE,Tapia2012CDC}, or principle orthogonal decompositions\cite{Vo2019}.
% XXX - would be good to look for a few other examples outside of our own papers where the CME dimension is reduced by linear projections.
On the other hand, an experimental obstacle to using CME analyses is that gene expression quantification based on fluorescence intensities (e.g., using fluorescent protein reporters) usually does not provide molecular resolution, but rather measures only relative changes in expression over time.

@@

\subsection{Introduction of Observation/Actuation Time Delays}
A time-delay SSA was developed by recording the piecewise constant history of the system and inputing the current control activation at time $t$ in the following delayed manner
\begin{equation}
u_{\tau}(t)=\left\{
\begin{array}{rl}
      0 ,&\text{ for }  t \leq \tau, \\
      {u}^{\mathcal{C}}(x_1(t-\tau), x_2(t-\tau)) , &\text{ for }   t > \tau,\\
\end{array}\right. 
\label{scoreeq}
\end{equation}
where $\tau$ is the time delay between observation and actuation, and ${u}^{\mathcal{C}}$ is the previously optimized control law for $\mathbf{u}^{FAC}$ or $\mathbf{u}^{PAC}$. We note that the time delay stochastic process was only simulated using the SSA because an appropriate direct FSP/CME integration procedure is not known.

\subsection{Parameter Scalings and Sensitivity}
Parameter sensitivities of the system can be performed by calculating the control performance score ($J$), over a range of parameters which are scaled linearly with a scaling parameter ($s$) while holding all other parameters constant. Any new arbitrary parameter $\lambda\in \{ \kappa,\beta,k_0,\eta,\gamma\}$ can be rescaled from an old one ($\lambda$) according to
\begin{equation}
\lambda' = s \lambda
\end{equation}
It should be noted that each cell receives its own set of parameters, and therefore the number of parameters double during sensitivity analysis of a two cell control process. Two types of sensitivity analysis can be performed, one which performs the analysis on one cell at a time while keeping the other constant, and another which scales similar parameters in both cells at the same time. Steady state distributions of the perrtubed models were solved using the FSP.
\section{Results}
\subsection{Effects of Changes to Granularity}
\begin{figure*}[t!]
\begin{center}
\includegraphics[width=1\textwidth]{GranularityPerturbation.png}
\vspace{-0.1in}
\caption{Effects of Scale Changes. (Ai-Fi) Joint (Ai,Ci,Di) and marginal (Bi,Di,Fi) probability distributions for the FAC control method at three different $\alpha$. $\alpha=0.2$ (top), $\alpha=1$ (middle) and $\alpha=2.0$ (bottom) relative to the target point (blue circles). (Aii-Fii) Same as (Ai-Fi), but for PAC controller. (G)  FAC (cyan) and PAC (magenta) controller performance as function of system volume. Dashed lines correspond to SSA analyses at $\alpha=100$. In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021} other than their adaptation to the new system volume size.}
\label{Volume}
\end{center}
\vspace{-0.2in}
\end{figure*}
Despite noise being a critical component of noise-exploiting control, the removal of noise through the addition of system granularity is likely beneficial to control performance. This two-edged characteristic that noise is critically important yet bad was explored by observing how control performance changes with system granularity, a parameter that is known to make systems behave more deterministically.

Control performance scores for the $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ controllers were calculated at different levels of granularity ($\alpha$) between 0.1 and 2.0. Figures.\ \ref{Volume} (Ai - Fi) show the joint (left plots) and marginal (right plots) distributions of the two cells relative to the specified target position (circles in joint distributions and vertical dashed lines in marginal distributions)
for the FAC controller and for $\mathbf{\alpha}$=.2 (A,B), 1 (C,D) and 2 (E,F). As the granularity changes from $\alpha$=0.2 , to $\alpha$=2, the distributions become focused at a point near the target, and the performance score improves from $J=$333 to 85 to 27, respectively. The PAC controller performance also improves considerably with alpha as shown in figures. \ref{Volume} (Aii - Fii).

Figure \ \ref{Volume}G shows the trend of the performance score versus alpha for both the FAC (solid cyan line) and the PAC (solid magenta line) controller. This improvement in performance appears to approach a small value as the granularity goes to infinity, but since the size of the FSP increases with the square of the system size, systems much larger than $\alpha$=2 (where $\mathbf{A}_0\in \mathbb{R}^{10^4\times10^4}$) are difficult to calculate. To bypass this limit in the FSP, sixteen SSA simulations were used to sample the CME of a system with a much larger volume of $\alpha$=100. Each SSA was ran for 5e7 minutes and only the last 4e7 minutes were sampled from. The performance score estimates of this high granularity SSA using the FAC and PAC were 4.13 and 27.5 respectively, which are plotted as dashed lines in figure \ref{Volume}G. We note however, that systems with larger granularity take longer to reach a steady state distribution, and it is unclear if performance improvements can be obtained with further increases to the system volume.  However, for all volumes considered, we found that both controllers monotonically improved with increased volume and that $\mathbf{u}^{FAC}$ always outperforms the $\mathbf{u}^{PAC}$ at any system volume.

These data suggest that increasing granularity increases steady state control performance even if the controller itself depends on noise for its implementation. Also, designing a controller to work for a system of one volume size can result in a controller that works even for another much larger volume.  This has practical significance, because it is relatively easy to search over a control law defined on a small $M\times M$ grid of states (e.g., when $M$=50 as in \cite{May2021}), but to implement the controller for a larger system which may be computationally prohibitive. This result implies that it may be possible to optimize controller using coarse-grained FSP analyses for small granularity problems and adapt these via interpolation for use with larger, more realistic systems, that exceed the computational limit of standard FSP computations.

@@

Time delay was hypothesized to adversely affect control performance of the $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ controllers. To test the effect of time delay on control performance, groups of time delayed SSA were simulated while changing controller and the level of time delay between each group. Each SSA was sampled for 1000 times over 10000 minutes after reaching a steady state by simulating the system for 10000 minutes.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=1\textwidth]{ParameterPerturbation.png}
\caption{Effects of parametric variations to Cell 1 (top row), Cell 2 (middle row) or to both cells simultaneously (bottom row). Each column show the effect of changing a single model parameter (e.g., $\kappa$ in the leftmost row). Trends for the $FAC$ (solid lines) and PAC (dashed lines), but the optimal controller can change under parameter variations. In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021}.}
\label{Parameter}
\end{center}
\end{figure*}

@@

It is rare that any parameters are exactly known for biological processes, and real systems often exhibit slightly different parameters due to extrinsic variations. This parameter variation between cells is factor that could influence control performance. To test this, parameter sensitivity analysis was performed on the cell one while holding the parameters of cell two constant figure \ref{Parameter}(Row I). For each perturbation analysis, both the $\mathbf{u}^{FAC}$ and $\mathbf{u}^{FAC}$ controller was used over a range of $s$ between 0.5 and 2. The data for such perturbations are shown in figure \ref{Parameter}. The analysis was then repeated, this time holding the parameters of Cell 1 constant and perturbing only the parameters of Cell 2  (shown in figure\ \ref{Parameter}(Row II)). Finally, both cell one and cell two had their parameters jointly perturbed by the same amount (figure \ref{Parameter}(Row III)).

Parameter changes were found to have a broad range of effects, including increasing performance, decreasing performance, or having little effect. Figure \ref{Parameter}(Bi) shows that increasing $\beta$ in Cell 1 worsens performance while increasing $\beta$ in Cell 2 improves performance. In some cases, these effects were not monotonic; for example increasing $\kappa$ in Cell 1 is highly advantageous up to a limit after which the control performance degrades rapidly. In other cases (such as for $k_0$), the effect of parameter perturbations on performance is insignificant even for relatively large ($s$=2) perturbations. When parameters of both Cell 1 and Cell 2 are jointly changed (Fig.\ \ref{Parameter} bottom row) some unique trends were observed. It was found that the $\mathbf{u}^{FAC}$ often outperformed the $\mathbf{u}^{PAC}$, but for large $s$ in a few parameters the inverse was found (figure \ref{parameter} (Row III, E and At).

When examining the effects of changing parameters when using the $\mathbf{u}^{FAC}$ (solid lines) or the $\mathbf{u}^{PAC}$ (dashed lines), general trends typically remained the same in that changes to a parameter which caused a decrease in the control performance of the FAC also tended to decrease the performance of the PAC. Some parameters (e.g., $\kappa$, $\gamma$) reached minima that performed better than the original, which suggests that the physical system itself has room for optimization beyond the original design that could lead to better control performance, {\em even without changing the control law}.  One explanation for why changes to $\eta$ had no effect on control performance is that the origonal parameter was so small that it may require a larger scaling on a different magnitude before changes are observed. For situations where the $\mathbf{u}^{PAC}$ outperformed the $\mathbf{u}^{FAC}$ at large $s$, this may be an indication that re-optimizing the controller for this parameter combination could yield better control results, since the controller is not re-optimized at each parameter combination.

>>>>

In this paper, the analysis of this Single-Input-Multiple-Output (SIMO) multicellular control problem is extended to examine how the FAC and PAC controllers are affected by additional model uncertainties, including coarse grained approximations to the system dynamics, uncertainties or errors in system parameters, and delays in the biochemical sensors or optogenetic controllers through which the external feedback is implemented. In the next section, we briefly describe our methods to formulate the chemical master equation to analyze multiple individual cells in the same spatially homogeneous environment; we define the FAC and PAC controllers and their effects the dynamics of cells within that environment; and we explain how the control law is optimized to achieve desired performance objectives. Next, in the `Results' section, we explore how granularity of the model formulation, inaccuracy of parametric assumptions, or introduction of time delays would affect the FAC and PAC control performance. Finally, in `Conclusions' we summarize our findings and discuss the implications that these controllers, and their robustness to modeling errors, may have on future systems in synthetic biology.

@@

More specifically, the open-loop infinitesimal generator, $\mathbf{A}_0$, is constructed according to:
\begin{equation}
[\mathbf{A}_0]_{ij} = \left\{
\begin{array}{rl}
-\sum_{\mu =1}^{2N_{\rm c}} w_{\mu}(\mathbf{X}_j), &\text{for }i=j,\\
w_{\mu}(\mathbf{X}_j), &\text{for }\mathbf{X}_i = \mathbf{X}_j+\mathbf{s}_\mu\\
0, & \text{otherwise,}
\end{array}\right. 
\label{InfA}
\end{equation}
and the feedback infinitesimal generator, $\mathbf{Bu}^{\mathcal{C}}$, of the controller is constructed according to
\begin{equation}
[\mathbf{Bu^{\mathcal{C}}]}_{ij} = \left\{
\begin{array}{rl}
- N_{\rm c}{u}^{\mathcal{C}}(\mathbf{X}_j), &\text{for }i=j\\
{u}^{\mathcal{C}}(\mathbf{X}_j), &\text{for }\begin{array}{ll}\mathbf{X}_i =\mathbf{X}_j + \mathbf{e}_{i_{\rm c}}, \\ \text{and } i_{\rm c} = 1,\ldots,N_{\rm c}\end{array},\\
0, & \text{otherwise,}
\end{array}\right.
\label{InfB}
\end{equation}
where ${u}^{\mathcal{C}}(\mathbf{X}_j)$ is the specification of the $\mathcal{C}$ controller in terms of the state, or its partial observations.

@@

\subsection{Scaling for system granularity}
One of the major computational obstacles in using the CME or FSP formulations to analyze discrete stochastic chemical kinetics is that the dimension of the FSP scales linearly with each chemical species, and therefore exponentially with the number of independent species. To circumvent this issue, several previous studies have projected the FSP onto lower dimensional spaces based using various techniques such as time scale separations\cite{Peles2006}, Krylov subspaces\cite{Macnemara&Sidje?}, coarse meshes\cite{Munsky2008IEEE,Tapia2012CDC}, or principle orthogonal decompositions\cite{Vo2019}.
% XXX - would be good to look for a few other examples outside of our own papers where the CME dimension is reduced by linear projections.
On the other hand, an experimental obstacle to using CME analyses is that gene expression quantification based on fluorescence intensities (e.g., using fluorescent protein reporters) usually does not provide molecular resolution, but rather measures only relative changes in expression over time.

@@

\subsection{Observation and actuation time delays}
No real optogenetic control system can provide instantaneous feedback -- there will always be delays in biochemical reactions needed for observations (e.g., transcription and translation dynamics as well as folding and maturation of fluorescent proteins), delays in data analysis and decision making (e.g., image processing and calculation of the corresponding control signals), and delays in the actuation dynamics (e.g., activation of optogenetically excitable molecules).  To explore the impacts of delays between the time a cellular fluctuation occurs and the time at which the corresponding control action can take effect, we developed a simple time-delay stochastic simulation algorithm. In this analysis, the state history is recorded following every stochastic event, allowing for reconstruction of the piecewise constant history of the species' populations, e.g., $x_1(s)$ and $x_2(s)$ for all $s\in [0,t]$.  Using this information, the time-delayed control signal at time $t$ can be specified as: 
\begin{equation}
u_{\tau}(t)=\left\{
\begin{array}{rl}
      0 ,&\text{ for }  t \leq \tau, \\
      {u}^{\mathcal{C}}(x_1(t-\tau), x_2(t-\tau)) , &\text{ for }   t > \tau,\\
\end{array}\right. 
\label{timeDelaySSA}
\end{equation}
where $\tau$ is the time delay between observation and actuation, and ${u}^{\mathcal{C}}$ is the previously optimized control law (e.g., $\mathbf{u}^{\rm FAC}$ or $\mathbf{u}^{\rm PAC}$). We note that the time delay stochastic process was only simulated using the SSA because to our knowledge an appropriate direct FSP/CME integration procedure has not yet been developed.

%\subsection{Parameter Scalings and Sensitivity}
%Finally, to explore the sensitivity of the control performance to inaccuracies in model parameters for one or more cells, we calculate the control performance score ($J$), over a range for each parameter scaled linearly with a scaling parameter ($\zeta$) while holding all other parameters constant. 
%%Any new arbitrary parameter $\lambda\in \{ \kappa,\beta,k_0,\eta,\gamma\}$ can be rescaled from an old one ($\lambda$) according to
%%\begin{equation}
%%\lambda' = s \lambda
%%\end{equation}
%We explore two types of sensitivity analysis: one which performs the analysis on one cell at a time while keeping the other constant, and another which scales similar parameters in both cells at the same time. 
%%In the former case, it should be noted that each cell receives its own set of parameters, so that the number of parameters double during sensitivity analysis of a two cell control process. 
%In either case, steady state distributions of the perturbed models are solved using the FSP.
% XXX - I don't think that we need this section in Methods -- but perhaps some of the text should be moved to the results if needed.

\section{Results}

To assess the potential robustness of noise-enhanced control of the proposed synthetic auto-repression toggle switch, we adopt the model and  parameters from \cite{May2021} as well as the fully aware and partially aware controllers, $u^{\rm FAC}(x_1,x_2)$ and  $u^{\rm FAC}(x_1)$, respectively as described in the methods section.  In the following subsections, we explore how well these controllers would perform in the more realistic setting where the approximate model used to define the controller no longer matches exactly to the dynamics of the process that the controller is being used to manipulate.

We first explore how the model defined for a given assumed system size (i.e., in which protein numbers have a range of $M$) when applied to a system with larger or smaller systems sizes (i.e., where protein numbers now have a range of $M' = \alpha M$.  This modification could arise from two situations related to numerical convenience or inexact resolution in the measurement of concentrations. First, from the perspective of numerical convenience when optimizing a controller, it can be advantageous to reduce the assumed range of protein levels in order to reduce the dimension of the CME and FSP analyses, thereby achieving a more tractable optimization problem.  For example, the current model (see Methods) takes XXX seconds to solve the FSP analysis when $\alpha=1$ but the same analysis takes XXX seconds to solve when $\alpha=10$. Since numerical optimization of the controller can take many millions of calculation for different control laws, learning a controller using $\alpha=1$ that works for $\alpha=10$ would yield substantial savings in computational effort. The second situation under which sensitivity to system scaling could become important is when measurements are relative and do not provide single-molecule resolution. For example, in measurements like those on \cite{Baumschlager} which are based on fluorescence protein intensities, it is not clear how many proteins correspond to what level of fluorescence intensity, and the scalar $\alpha$ is an unknown quantity that must be estimated (e.g., through calibration with single-molecule measurements or dilution experiment\cite{XXX - look for elowitz paper or something similar that tried to calibrate for single-molecule counts}).

It is well established that the relative size of stochastic fluctuations (i.e., noise) in a chemical process decreases with the inverse square root of the process scale \cite{XXX van kampen? Gillespei look for this}. As the system size increases (with all species being scaled by the same factor), the dynamics approach those of a deterministic process with probability one (with zero-measure exceptions only for initial conditions lying on manifolds that separate attractors for different steady state behaviors)\cite{XXX}. For the purpose of noise-induced control, this introduces a tradeoff where noise is critical to break symmetry and enable differential control, yet noise is also deleterious to maintaining desired states once they have been achieved. Considering these competing objectives, it is not immediately clear how the reduction of noise through the addition of system granularity will affect control performance.

To shed light on this tradeoff, control performance scores for the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers were calculated at different levels of granularity ($\alpha$) between 0.1 and 2.0. 
% XXX - Why only go up to 2?  Did it become intractable?
Figure \ref{Volume} (Ai - Fi) shows the joint (left plots) and marginal (right plots) distributions of the two cells relative to the specified target position (circles in joint distributions and vertical dashed lines in marginal distributions) for the FAC controller and for $\mathbf{\alpha}$=.2 (A,B), 1 (C,D) and 2 (E,F). As the granularity changes from $\alpha$=0.2 , to $\alpha$=2, the distributions become focused at a point near the target, and the performance score improves from $J=$333 to 85 to 27, respectively. The PAC controller performance also improves considerably with alpha as shown in Fig.\ \ref{Volume} (G-L).
% XXX - Please change the panel labels to A-F, G-L, and M.

\subsection{Effects of changes to system granularity}
\begin{figure*}[t!]
\begin{center}
\includegraphics[width=1\textwidth]{GranularityPerturbation.png}
\vspace{-0.1in}
\caption{Effects of Scale Changes. (A-F) Joint (A,C,D,) and marginal (B,D,F) probability distributions for the FAC control method at $\alpha=0.2$ (A,B), $\alpha=1$ (C,D) and $\alpha=2.0$ (E,F).  Target point shown as blue circle. (G-L) Same as (A-F), but for PAC controller. (M)  FAC (cyan) and PAC (magenta) controller performance as function of system volume. Horizontal dashed lines correspond to SSA analyses at $\alpha=100$. In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021} other than their adaptation to the new system volume size (see Methods).}
\label{Volume}
\end{center}
\vspace{-0.2in}
\end{figure*}

Figure \ref{Volume}M shows the trend of the performance score versus alpha for both the FAC (solid cyan line) and the PAC (solid magenta line) controller. This improvement in performance appears to approach a small value as the granularity goes to infinity, but since the size of the FSP increases with the square of the system size, systems much larger than $\alpha$=2 (where $\mathbf{A}_0\in \mathbb{R}^{10^4\times10^4}$) become more difficult to calculate. 
% XXX - I would think you should be able to compute larger values of 10^6.  They would be slow, but should be doable in a few minutes per calculation.
To bypass this limit in the FSP, sixteen SSA simulations were used to sample the CME of a system with a much larger volume of $\alpha$=100. Each SSA was ran for $5\times10^7$ minutes and only the last $4\times10^7$ minutes were sampled to estimate the stationary distribution and the calculate the performance score. The performance score estimates of this high granularity SSA using the FAC and PAC were 4.13 and 27.5 respectively, which are plotted as dashed lines in Fig.\ \ref{Volume}G. Although it is unclear if further performance improvements could be obtained with further increases to the system scale, for all cases considered so far, we found that both controllers monotonically improved with increased $\alpha$ and that $\mathbf{u}^{\rm FAC}$ always outperforms the $\mathbf{u}^{\rm PAC}$.

We note however, that the improved steady state performance for systems with larger scales, comes at a cost that they also take longer to reach a steady state distribution. This is illustrated in Fig.\ \ref{XXX} which shows the time needed to adapt to a change in the target point $\mathbf{T} = [XXX,XXX]$ to its mirror point $\mathbf{T}_m = [XXX,XXX]$.  
% XXX - Please add a new figure to show this point.  Start with an initial condition that is the steady state with target T and then flip it to T' = T(::-1).  Then solve for P(t) and calculated J(t) from this.    Finally, make a plot of J(t) versus t for the three cases \alpha = 0.2, 1, 2.
 From the figure, it is clear that the smaller systems XXX.
 % XXX - update this to explain what you see in this new figure.

%These data suggest that increasing granularity increases steady state control performance even if the controller itself depends on noise for its implementation. Also, designing a controller to work for a system of one volume size can result in a controller that works even for another much larger volume.  This has practical significance, because it is relatively easy to search over a control law defined on a small $M\times M$ grid of states (e.g., when $M$=50 as in \cite{May2021}), but to implement the controller for a larger system which may be computationally prohibitive. This result implies that it may be possible to optimize controller using coarse-grained FSP analyses for small granularity problems and adapt these via interpolation for use with larger, more realistic systems, that exceed the computational limit of standard FSP computations.

@@

Time delay was hypothesized to adversely affect control performance of the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controllers. To test the effect of time delay on control performance, groups of time delayed SSA were simulated while changing controller and the level of time delay between each group. Each SSA was sampled for 1000 times over 10000 minutes after reaching a steady state by simulating the system for 10000 minutes.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=1\textwidth]{ParameterPerturbation.png}
\caption{Effects of parametric variations to Cell 1 (top row), Cell 2 (middle row) or to both cells simultaneously (bottom row). Each column show the effect of changing a single model parameter (e.g., $\kappa$ in the leftmost row). Trends for the FAC (solid lines) and PAC (dashed lines), but the optimal controller can change under parameter variations. In all analyses, the system model and PAC and FAC controllers are unchanged from \cite{May2021}.}
\label{Parameter}
\end{center}
\end{figure*}

@@

It is rare that any parameters are exactly known for biological processes, and real systems often exhibit slightly different parameters due to extrinsic variations. This parameter variation between cells is factor that could influence control performance. To test this, parameter sensitivity analysis was performed on the cell one while holding the parameters of cell two constant figure \ref{Parameter}(Row I). For each perturbation analysis, both the $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ controller was used over a range of $s$ between 0.5 and 2. The data for such perturbations are shown in figure \ref{Parameter}. The analysis was then repeated, this time holding the parameters of Cell 1 constant and perturbing only the parameters of Cell 2  (shown in figure\ \ref{Parameter}(Row II)). Finally, both cell one and cell two had their parameters jointly perturbed by the same amount (figure \ref{Parameter}(Row III)).

Parameter changes were found to have a broad range of effects, including increasing performance, decreasing performance, or having little effect. Figure \ref{Parameter}(Bi) shows that increasing $\beta$ in Cell 1 worsens performance while increasing $\beta$ in Cell 2 improves performance. In some cases, these effects were not monotonic; for example increasing $\kappa$ in Cell 1 is highly advantageous up to a limit after which the control performance degrades rapidly. In other cases (such as for $k_0$), the effect of parameter perturbations on performance is insignificant even for relatively large ($s$=2) perturbations. When parameters of both Cell 1 and Cell 2 are jointly changed (Fig.\ \ref{Parameter} bottom row) some unique trends were observed. It was found that the $\mathbf{u}^{\rm FAC}$ often outperformed the $\mathbf{u}^{\rm PAC}$, but for large $s$ in a few parameters the inverse was found (figure \ref{parameter} (Row III, E and At).

When examining the effects of changing parameters when using the $\mathbf{u}^{\rm FAC}$ (solid lines) or the $\mathbf{u}^{\rm PAC}$ (dashed lines), general trends typically remained the same in that changes to a parameter which caused a decrease in the control performance of the FAC also tended to decrease the performance of the PAC. Some parameters (e.g., $\kappa$, $\gamma$) reached minima that performed better than the original, which suggests that the physical system itself has room for optimization beyond the original design that could lead to better control performance, {\em even without changing the control law}.  One explanation for why changes to $\eta$ had no effect on control performance is that the origonal parameter was so small that it may require a larger scaling on a different magnitude before changes are observed. For situations where the $\mathbf{u}^{\rm PAC}$ outperformed the $\mathbf{u}^{\rm FAC}$ at large $s$, this may be an indication that re-optimizing the controller for this parameter combination could yield better control results, since the controller is not re-optimized at each parameter combination.
